---
title: "Fast multi-scale neighbour embedding"
author:
  - name: Laurent Gatto
    affiliation:
    - Computational Biology and Bioinformatics, UCLouvain
    email: laurent.gatto@uclouvain.be
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
bibliography: fmsne.bib
date: "`r doc_date()`"
package: "`r pkg_ver('fmsne')`"
vignette: >
  %\VignetteIndexEntry{The fmsne package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
    )
suppressPackageStartupMessages(library("BiocStyle"))
```


# Introduction

## The `fmsne` package

The `fmsne` R package implements the [fast multi-scale neighbour
embedding](https://github.com/lgatto/Fast_Multi-scale_NE) methods
developed by [Cyril de Bodt](https://github.com/cdebodt).

The fast algorithms which are implemented are described in the article
[Fast Multiscale Neighbor
Embedding](https://ieeexplore.ieee.org/document/9308987), from Cyril
de Bodt, Dounia Mulders, Michel Verleysen and John A. Lee, published
in IEEE Transactions on Neural Networks and Learning Systems, in 2020.

The implementations are provided using the python programming
language, but involve some C and Cython codes for performance
purposes.

If you use the codes in this repository or the article, please cite as
[@De_Bodt2022-1]:

- C. de Bodt, D. Mulders, M. Verleysen and J. A. Lee, "Fast Multiscale
  Neighbor Embedding," in IEEE Transactions on Neural Networks and
  Learning Systems, 2020, doi: 10.1109/TNNLS.2020.3042807.

and this package:

```{r}
citation("fmsne")
```

## Installation instructions

To install this R package, simply run

```r
BiocManager::install("lgatto/fmsne")
```

The package on the following Bioconductor packages:

- [SingleCellExperiment](https://bioconductor.org/packages/SingleCellExperiment)
  for the infrastructure to hold the single-cell and reduced dimension
  data.

- [basilisk](https://bioconductor.org/packages/basilisk) to install
  and run the underlying Python implementation.


If you are looking to apply fast multi-scale neighbor embedding in
Pyhton, you can install the `fmsne` python package with

```r
pip install fmsne
```

# Package functionality

## Neighbor Embedding

- `runMSSNE()`: nonlinear dimensionality reduction through multi-scale
  (MS) stochastic neighbor embedding (SNE) [@Van_der_Maaten2008-7;
  @Van_Der_Maaten2014-8], as presented in the reference [@Lee2015-2]
  below and summarized in [@De_Bodt2022-1].

- `runMSTSNE()`: nonlinear dimensionality reduction through
  multi-scale t-distributed SNE (t-SNE) [@Van_der_Maaten2008-7;
  @Van_Der_Maaten2014-8], as presented in the reference
  [@De_Bodt_undated-6] below and summarized in [@De_Bodt2022-1].

- `runFMSSNE()`: nonlinear dimensionality reduction through fast
  multi-scale SNE (FMS SNE), as presented in the reference
  [@De_Bodt2022-1].

- `runFMSTSNE()`: nonlinear dimensionality reduction through fast
  multi-scale t-SNE (FMS t-SNE), as presented in the reference
  [@De_Bodt2022-1].

See the function manual pages for further details

## Quality control

- `drQuality()`: unsupervised evaluation of the quality of a
  low-dimensional embedding, as introduced in [@Lee2009-3; @Lee2010-4]
  and applied and summarized in [@De_Bodt2022-1; @Lee2015-2;
  @Lee2013-5]. This function assesses the dimensionality reduction
  quality measuring the neighborhood preservation from the
  high-dimensional space to the low-dimensional one. The documentation
  of the function explains the meaning of the criteria and how to
  interpret them.

# Session information

```{r si, echo = FALSE}
sessionInfo()
```

# References
