---
title: "Fast multi-scale neighbour embedding"
author:
  - name: Laurent Gatto
    affiliation:
    - Computational Biology and Bioinformatics, UCLouvain
    email: laurent.gatto@uclouvain.be
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('fmsne')`"
vignette: >
  %\VignetteIndexEntry{The fmsne package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
    )
suppressPackageStartupMessages(library("BiocStyle"))
```


# Introduction

## The `fmsne` package

The `fmsne` R package implements the [fast multi-scale neighbour
embedding](https://github.com/lgatto/Fast_Multi-scale_NE) methods
developed by [Cyril de Bodt](https://github.com/cdebodt).


The fast algorithms which are implemented are described in the article
[Fast Multiscale Neighbor
Embedding](https://ieeexplore.ieee.org/document/9308987), from Cyril
de Bodt, Dounia Mulders, Michel Verleysen and John A. Lee, published
in IEEE Transactions on Neural Networks and Learning Systems, in 2020.

The implementations are provided using the python programming
language, but involve some C and Cython codes for performance
purposes.

If you use the codes in this repository or the article, please cite
as:

> C. de Bodt, D. Mulders, M. Verleysen and J. A. Lee, "Fast Multiscale
> Neighbor Embedding," in IEEE Transactions on Neural Networks and
> Learning Systems, 2020, doi: 10.1109/TNNLS.2020.3042807.


## Installation instructions

To install this R package, simply run

```r
BiocManager::install("lgatto/fmsne")
```

The package on the following Bioconductor packages:

- [SingleCellExperiment](https://bioconductor.org/packages/SingleCellExperiment)
  for the infrastructure to hold the single-cell and reduced dimension
  data.

- [basilisk](https://bioconductor.org/packages/basilisk) to install
  and run the underlying Python implementation.


If you are looking to apply fast multi-scale neighbor embedding in
Pyhton, you can install the `fmsne` python package with

```r
pip install fmsne
```

# Package functionality

Neighbor Embedding

- `mssne`: nonlinear dimensionality reduction through multi-scale SNE
  (Ms SNE), as presented in the reference [2] below and summarized in
  [1]. This function enables reducing the dimension of a data
  set. Given a data set with N samples, the 'mssne' function has
  O(N**2 log(N)) time complexity.

- `mstsne`: nonlinear dimensionality reduction through multi-scale
  t-SNE (Ms t-SNE), as presented in the reference [6] below and
  summarized in [1]. This function enables reducing the dimension of a
  data set. Given a data set with N samples, the 'mstsne' function has
  O(N**2 log(N)) time complexity.

- `fmssne`: nonlinear dimensionality reduction through fast
  multi-scale SNE (FMs SNE), as presented in the reference [1]
  below. This function enables reducing the dimension of a data
  set. Given a data set with N samples, the 'fmssne' function has O(N
  (log(N))**2)

- `fmstsne`: nonlinear dimensionality reduction through fast
  multi-scale t-SNE (FMs t-SNE), as presented in the reference [1]
  below. This function enables reducing the dimension of a data
  set. Given a data set with N samples, the 'fmstsne' function has O(N
  (log(N))**2) time complexity.

Quality control

- `drQuality`: unsupervised evaluation of the quality of a
  low-dimensional embedding, as introduced in [3, 4] and employed and
  summarized in [1, 2, 5]. This function enables computing DR quality
  assessment criteria measuring the neighborhood preservation from the
  high-dimensional space to the low-dimensional one. The documentation
  of the function explains the meaning of the criteria and how to
  interpret them. Given a data set with N samples, the
  'eval_dr_quality' function has O(N**2 log(N)) time complexity. It
  can hence run using databases with up to a few thousands of
  samples. This function is not based on the Cython implementations in
  `fmsne_implem.pyx`.

<!-- - `red_rnx_auc`: this function is similar to the `eval_dr_quality` -->
<!--   function, but given a data set with N samples, the `red_rnx_auc` -->
<!--   function has O(N*Kup*log(N)) time complexity, where Kup is the -->
<!--   maximum neighborhood size accounted when computing the quality -->
<!--   criteria. This function can hence run using much larger databases -->
<!--   than `eval_dr_quality`, provided that Kup is small compared to -->
<!--   N. This function is based on the Cython implementations in -->
<!--   `fmsne_implem.pyx`. -->

Notations

- DR: dimensionality reduction.
- HD: high-dimensional.
- LD: low-dimensional.
- HDS: HD space.
- LDS: LD space.
- SNE: stochastic neighbor embedding.
- t-SNE: t-distributed SNE.
- Ms SNE: multi-scale SNE.
- Ms t-SNE: multi-scale t-SNE.
- BH t-SNE: Barnes-Hut t-SNE.


## References

[1] C. de Bodt, D. Mulders, M. Verleysen and J. A. Lee, "Fast
Multiscale Neighbor Embedding," in IEEE Transactions on Neural
Networks and Learning Systems, 2020, doi: 10.1109/TNNLS.2020.3042807.

[2] Lee, J. A., Peluffo-Ordóñez, D. H., & Verleysen,
M. (2015). Multi-scale similarities in stochastic neighbour embedding:
Reducing dimensionality while preserving both local and global
structure. Neurocomputing, 169, 246-261.

[3] Lee, J. A., & Verleysen, M. (2009). Quality assessment of
dimensionality reduction: Rank-based criteria. Neurocomputing,
72(7-9), 1431-1443.

[4] Lee, J. A., & Verleysen, M. (2010). Scale-independent quality
criteria for dimensionality reduction. Pattern Recognition Letters,
31(14), 2248-2257.

[5] Lee, J. A., Renard, E., Bernard, G., Dupont, P., & Verleysen,
M. (2013). Type 1 and 2 mixtures of Kullback–Leibler divergences as
cost functions in dimensionality reduction based on similarity
preservation. Neurocomputing, 112, 92-108.

[6] de Bodt, C., Mulders, D., Verleysen, M., & Lee,
J. A. (2018). Perplexity-free t-SNE and twice Student tt-SNE. In ESANN
(pp. 123-128).

[7] van der Maaten, L., & Hinton, G. (2008). Visualizing data using
t-SNE. Journal of Machine Learning Research, 9(Nov), 2579-2605.

[8] van der Maaten, L. (2014). Accelerating t-SNE using tree-based
algorithms. Journal of Machine Learning Research, 15(1), 3221-3245.

# Session information

```{r si, echo = FALSE}
sessionInfo()
```
